Seniority and hierarchy configuration interaction for radicals and excited states

### 2023-08-06 Manuscript/manuscript.tex
saving work

% Abstract
\begin{abstract}
{\bf Abstract:} 
%\bigskip
%\begin{center}
%        \boxed{\includegraphics[keepaspectratio,width=2in]{TOC}}
%\end{center}
%\bigskip
\end{abstract}

@@

The most well-known CI route is excitation-based.
Starting from a reference determinant, typically the Hartree-Fock (HF) determinant,
one generates all determinants that are connected to it by excitating at most $e$ electrons.
The excitation degree $e$ thus defines the order of the approximate CI model:
CI with single excitations (CIS); CI with single and double excitations (CISD), CI with single, double and triple excitations (CISDT), and so on.
A different CI route is based on the seniority number $s$ (the number of unpaired electrons in a given determinant).
In seniority-based CI (sCI) \cite{Bytautas_2011,Allen_1962,Smith_1965,Veillard_1967}, there is no reference determinant, and one accounts for all determinants having seniority equal or less than $s$.
For systems with an even number of electrons, the first approximate model is defined by $s=0$ (sCI0), usually refered to as doubly-occupied CI (DOCI),
which is followed by the higher-order models sCI2, sCI4, and so on.
For an odd number of electrons, the seniority route follows along odd numbers of $s$: sCI1, sCI3, and so on.
%Limited studies for radicals and excited states

%sCI works well to describe molecular dissociation \cite{Bytautas_2015,Alcoba_2014,Alcoba_2014b}
%For small systems, FCI is still attainable and one can thus gauge the performance of approximate CI models.
We have recently introduced a third CI route, hierarchy CI (hCI), \cite{Kossoski_2022}
where the Hilbert space is partitioned according to a hierarchy parameter $h$ that combines the excitation degree $e$ and the seniority number $s$ according to $h = (e+s/2)/2$.
This definition ensures that all types of determinants whose number share the same scaling with system size are included at the same hierarchy $h$.
For different properties and closed-shell molecular systems, 
hCI was found to display an overall faster convergence to the FCI results than the traditional excitation-based CI. \cite{Kossoski_2022}
However, hCI has only been defined for a closed-shell reference, thus systems with an even number of electrons.
Here, our first goal is to generalize hCI for open-shell systems and to assess the performance of this class of methods by performing calculations for various properties of several radicals.
%A generalization of the method for open-shell systems has not yet been proposed.
%excitations of increasingly higher order give rise to a systematically improvable sequence of CI methods:
The second goal is to evaluate how perturbation theory, more precisely the second-order Epstein-Nesbet (EN2) perturbative correction (both standard and renormalized), \cite{Garniron_2019}
impacts different properties of ground-state closed-shell systems and radicals, for excitation-based, sCI and hCI methods.

%Whereas the first CI root represents the electronic ground state, higher-lying roots give access to excited states.
%Similarly, the favourable results of hCI when compared to excitation-based CI for ground states raises the question on how well it can describe excitation energies.
%The performance of excitation-based CI for excited states is well stablished, \cite{} but remains an open question for sCI methods,
Inspired by the results of hCI for ground-state closed-shell systems, \cite{Kossoski_2022} our third goal is to define and assess the accuracy of hCI models for excited states.
Furthermore, to the best of our knowledge, sCI models have not yet been directly used to target excited states,
despite the growing number of methods that exploit the concept of seniority number,
for both ground
\cite{Limacher_2013,Limacher_2014,Tecmer_2014,Boguslawski_2014a,Boguslawski_2015,Boguslawski_2014b,Boguslawski_2014c,Johnson_2020,Henderson_2014,Stein_2014,Henderson_2015,Chen_2015,Bytautas_2018,Marie_2021,Boguslawski_2021,Tecmer_2022,Mamache_2023} 
and excited states.
\cite{Boguslawski_2016b,Boguslawski_2016c,Boguslawski_2019,Nowak_2019,Kossoski_2021,Marie_2021,Tecmer_2022,Rishi_2023,Nowak_2023} 
Our fourth goal is therefore to introduce and gauge sCI models for excited states.
There are two possible approaches to target excited states with CI.
One can employ the ground-state HF orbitals and obtain excitation energies from the higher-lying eigenvalues of the CI matrix, which we refer to as ground-state based approach.
Instead, one may optimize the orbitals for the excited state of interest (described with an appropriate reference), followed by a separate CI calculation,
in a so-called state-specific approach ($\Delta$CI).
There has been a recent surge in the development of state-specific methods, covering 
single-reference and multiconfigurational self-consistent field,
\cite{Ziegler_1977,Burton_2021,Shea_2018,Tran_2019,Tran_2020,Hardikar_2020,Burton_2022,Hanscam_2022,Kossoski_2023,Marie_2023}
% add references for SCF
%\cite{Ziegler_1977,Burton_2021}
%multiconfigurational SCF, \cite{Shea_2018,Tran_2019,Tran_2020,Hardikar_2020,Burton_2022,Kossoski_2023,Marie_2023}
density functional theory, 
\cite{Filatov_1999,Kowalczyk_2011,Kowalczyk_2013,Gilbert_2008,Barca_2018,Hait_2020,Hait_2021,Hardikar_2020,Zhao_2020,Levi_2020,Carter-Fenk_2020,Toffoli_2022,Schmerwitz_2022}
and coupled-cluster (CC)
\cite{Piecuch_2000,Mayhall_2010,Lee_2019,Kossoski_2021,Marie_2021,Rishi_2023}
methods.
In particular, by employing a minimal configuration state function (CSF) reference, 
we have recently shown that excitation-based $\Delta$CI models deliver far more accurate excitation energies than the ground-state based analogues. \cite{Kossoski_2023}
Based on this promising set of results for excitation-based CI, here we explore both ground-state based and state-specific approaches for sCI and hCI models.

@@

We introduced hCI \cite{Kossoski_2022} as a truncation of the Hilbert space based on a two-dimensional map of determinants 
based on their seniority $s$ and their excitation degree $e$ with respect to a reference determinant (see Fig.~\ref{fig:hCI}).
It is built such that all classes of determinants whose number $\Ndet$ share the same scaling with the number of basis functions $N$ enter at the same order of truncation,
which means moving diagonaly in this seniority-excitation map.

@@

hCI was initially defined for a closed-shell reference. \cite{Kossoski_2022}
Here we generalize hCI for an arbitrary type of reference Slater determinant and for systems with an odd number of electrons.
With respect to a given reference determinant, we define the hierarchy $h$ of a target determinant as
\begin{equation}
  \label{eq:h}
  h = \frac{e+ (s-s_0)/2}{2},
\end{equation}
where $s$ and $s_0$ denote the seniority of target and reference determinants, respectively, and $e$ represents the excitation degree that connects the two determinants.
The term $(s-s_0)/2$ is always an integer, with $s$ and $s_0$ being even (odd) for systems with an even (odd) numbers of electrons, whereas $e$ is also an integer.
Therefore, $h$ can assume integer or half-integer values, for any type of reference determinant.
For a given hCI model defined by $h$, we include all the determinants having a hierarchy less than or equal to $h$ with respect to any determinant in the reference.
Notice that Eq.~\eqref{eq:h} simplifies to the previous definition \cite{Kossoski_2022} for the case of a closed-shell reference determinant, when $s_0 = 0$.

The above definition ensures that all types of determinants with a given scaling enter at the same hierarchy $h$.
% computational scaling
We show in Fig.~\ref{fig:hCI} how the seniority-excitation map of determinants is partitioned for a closed-shell reference, and for open-shell references with one or two unpaired electrons.
For these three types of reference, Fig.~\ref{fig:determinants} shows the classes of determinants included up to hCI1.
hCI can also be interpreted as the CI model that provides increasingly more dissimilar determinants from the given reference determinant.
The level of dissimilarity is represented by the hierarchy $h$, which accounts for differences in orbital occupation (through the excitation degree $e$)
and differences in the number of unpaired electrons (through the term $(s-s_0)/2$).

%%% FIG 2 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{determinants}
\caption{Classes of determinants generated in hCI0, hCI0.5 and hCI1, for three different reference determinants:
closed-shell (top left), open-shell with one unpaired electron (top right) and open-shell with two unpaired electrons (bottom).
Each class is determined by the specific seniority and occupation of the orbital blocks of the reference (doubly-occupied, singly-occupied and unoccupied).
}
\label{fig:determinants}
\end{figure*}
%%% %%% %%%

The first order for a closed-shell reference determinant is hCI1 (see top left of Fig.~\ref{fig:determinants}), which accounts for all single excitations (as CIS) 
plus all paired doubled excitations (two electrons promoted from the same occupied orbital into the same virtual orbital).
For both types of determinants, their number scales as $N^2$, and are thus taken into account at the same hierarchy of hCI ($h=1$ in this case).
An odd number of excitations from a closed-shell reference always leave unpaired electrons, thus the empty blocks at $s=0$ and $e$ odd.
hCI1.5 includes to the hCI1 determinants the set of double excitations where two electrons are unpaired.
At the next integer order, hCI2 accounts for all types of determinants where $\Ndet$ scale as $N^4$ or less.
That means all single and double excitations (as CISD), plus the subset of triple excitations that leave only two unpaired electrons, 
plus the subset of quadruple excitations where no electrons are unpaired.
More generally, in hCI, the number of determinants $\Ndet$ scales polynomially with the number of basis functions $N$ as $N^{2h}$.
Similarly, for excitation-based CI, $\Ndet$ scales as $N^{2e}$, where $e$ is the excitation degree.
The computational scaling along the two routes is shown in Tab.~\ref{tab:scaling}.

\begin{table}[ht!]
\caption{Computational scaling along the hierarchy-based and excitation-base CI routes, with and without the EN2 perturbative correction.}
\label{tab:scaling}
\begin{ruledtabular}
\begin{tabular}{ll|ll|l}
 \mc{2}{c|}{hierarchy-based}   &  \mc{2}{c|}{excitation-based} &  scaling \\
\hline
hCI1   &    -       & CIS    &   -       & $\order*{N^4}$ \\
hCI1.5 &    -       &  -     &   -       & $\order*{N^5}$ \\
\hline
hCI2   & hCI1+EN2   & CISD   & CIS+EN2   & $\order*{N^6}$ \\
hCI2.5 & hCI1.5+EN2 &  -     &   -       & $\order*{N^7}$ \\
\hline
hCI3   & hCI2+EN2   & CISDT  & CISD+EN2  & $\order*{N^8}$ \\
hCI3.5 & hCI2.5+EN2 &   -    &   -       & $\order*{N^9}$ \\
\hline
hCI4   & hCI3+EN2   & CISDTQ & CISDT+EN2 & $\order*{N^{10}}$ \\
\end{tabular}
\end{ruledtabular}
\end{table}

For an odd number of electrons, $s$ assumes odd values starting from $s=1$.
The simplest reference is an open-shell determinant with one unpaired electron (see center pannel of Fig.~\ref{fig:hCI} and top right of Fig.~\ref{fig:determinants}).
In contrast to the closed-shell reference case, here there are no empty blocks in the excitation-seniority map.
Actually, the hCI series presents the hCI0.5 level, accounting for the single excitations from and into the singly-occupied orbital only.
For HF orbitals, however, these excitations do not connect with the reference, and hCI0.5 thus provides the same energy as restricted open-shell HF.
More generally, since there are more types of orbitals for the open-shell case (doubly-occupied, singly-occupied and unoccupied) than for the closed-shell case (doubly-occpupied or unoccupied),
there is correspondingly more classes of determinants at a given level of hCI.
This can be appreciated from the comparison of hCI1 for closed- and open-shell references, shown in Fig.~\ref{fig:determinants}.

Finally, one can employ a high-spin reference determinant to define the hCI models.
The case of two unpaired electrons is shown in the right pannel of Fig.~\ref{fig:hCI}.
The key difference with respect to the closed-shell case is that the hCI sequence is displaced one block to the right in the seniority-excitation map,
reflecting the shift from the $s_0=0$ to the $s_0=2$ reference.
In the two previous cases, the hCI0 level only includes the reference determinant.
For a $s=2$ reference, however, hCI0 not only accounts for the reference determinant,
but also the two closed-shell determinants produced by the single excitation that pairs the two unpaired electrons.
The lower order classes of determinants can be seen in the bottom of Fig.~\ref{fig:determinants},
which clearly outnumber the fewer classes associated with a closed-shell reference.
One could adopt even higher-spin references, further displacing the hCI sequence to the right in the seniority-excitation map.
In this case, hCI0 would include all the lower spin determinants obtained by partially or totally pairing the unpaired electrons.
Notice that the number of such determinants do not depend on the size of the system (scaling as $N^0$), and are thus included at the $h=0$ level, in line with the spirit of hCI.
Similarly, hCI can be built on top of a high-spin reference for odd number of electrons.

@@

The hCI models introduced here were implemented in {\QP} \cite{Garniron_2019} through a straightforward modification of the
\textit{configuration interaction using a perturbative selection made iteratively} (CIPSI) algorithm. \cite{Huron_1973,Giner_2013,Giner_2015,Garniron_2018}
By allowing only for the determinants that are connected with the reference determinant(s) up to a given maximum hierarchy $h$,
the CIPSI algorithm is restricted to the truncated Hilbert space defined by the reference determinant(s) and the value of $h$.
{\QP} \cite{Garniron_2019} was also employed to perform all the CI calculations presented here.
In a given calculation, the energies are considered to be converged when the (largest) EN2 correction computed in the truncated Hilbert space 
lies below \SI{0.01}{\milli\hartree}. \cite{Garniron_2018}
%by spanning the most important regions of the truncated Hilbert space
This selected CI procedure requires considerably fewer determinants than the total number of determinants in the truncated Hilbert space,
while delievering fairly converged absolute and excitation energies.
The ground- and excited-state CI energies are obtained with the Davidson iterative algorithm. \cite{Davidson_1975}
For a given approximate CI model, we futher evaluated the standard and the renormalized EN2 perturbative correction. \cite{Garniron_2019} 
This calculation involves a single loop over the determinants left outside the truncated (internal) space but connected to it via at most double excitations.
Looping over these external doubly excited determinants has a computational scaling equal to $\Ndet$,
thus $\order*{N^{2e+4}}$ for excitation-based CI and $\order*{N^{2h+4}}$ for hCI, where $e$ or $h$ define the internal CI space.
For example, hCI2 and CISD present a $\order*{N^6}$ computational scaling, whereas hCI2+EN2 and CISD+EN2 scale as $\order*{N^8}$,
though with a small prefactor stemming from the EN2 calculation.
The computational scaling associated with the CI+EN2 calculations is also presented in Tab.~\ref{tab:scaling}.

To gauge the performance of hCI against excitation-based and seniority-based CI for radical systems,
we have calculated the ground-state potential energy curves (PECs) for the dissociation of four radical systems:
\ce{OH}, \ce{CN}, vinyl (\ce{C2H3}), and \ce{H7}.
These calculations employed the ground-state restricted open-shell HF orbitals and the frozen core approximation.
The equilibrium geometry of vinyl was taken from Ref.~\onlinecite{Loos_2020},
and is also reproduced in the {\SupInf}.
Their PECs were computed along the \ce{C=C} double bond breaking coordinate, with the remaining internal coordinates kept frozen.
For \ce{H7}, we considered equally spaced and linearly arranged hydrogen atoms, and the PECs were computed along the symmetric dissociation coordinate.

The results were analyzed along the same lines of our report on hCI for closed-shell systems. \cite{Kossoski_2022}
Namely, for the different CI models considered here, 
we evaluated the convergence of the non-parallelity error (NPE), the distance error, the harmonic vibrational frequencies, and the equilibrium bond lengths, as functions of $\Ndet$.
The NPE of a given level of theory is defined as the maximum minus the minimum differences between its corresponding PEC and the FCI one, for a given range of coordinates.
Here we redefine the previous definition of the distance error \cite{Kossoski_2022} to accommodate for the fact the approximate PEC might appear below the FCI when perturbative corrections are employed.
The previous definition could provide negative values in this case, whereas this metric should provide the overall extent of proximity between two PECs.
Here, we define the distance error based on the signed differences between two PECs, as the absolute value of their maximum difference plus the absolute value of their minimum difference, 
evaluated at given coordinate interval.
In the discussion, we say equilibrium properties when referring to both the equilibrium geometry and the vibrational frequency.
Details about how the harmonic vibrational frequencies and equilibrium geometries were obtained from the calculated PECs,
along with the ranges defining the NPE and distance errors can be found in the \SupInf.

>>>>

% Abstract
\begin{abstract}
{\bf Abstract:}
Hierarchy configuration interaction (hCI) has been recently introduced as an alternative configuration interaction (CI) route combining excitation degree and seniority number, 
which showed to efficiently recover dynamic and static correlation for closed-shell molecular systems [J. Phys. Chem. Lett. 2022, 13, 19, 4342].
Here, we generalize hCI for arbitrary reference determinants, allowing calculations for radicals and for excited states in a state-specific approach.
We gauge this route against excitation-based CI (eCI) and seniority-based CI (sCI)
by evaluating how different properties of ground-state radicals converge to the full CI limit.
hCI outperforms or matches eCI, whereas sCI is far less accurate, in line with previous observations for closed-shell molecules.
Employing the second-order Epstein-Nesbet perturbation theory as a correction significantly accelerates the convergence of hCI and eCI.
We further explore various hCI and sCI models to calculate excitation energies of closed- and open-shell systems.
Our results underline that both the choice of reference and orbitals drive the fine balance between correlation of ground and excited states.
State-specific hCI2 and higher order models perform similarly to their eCI counterparts, whereas lower orders of hCI deliver poor results.
In turn, sCI1 produces decent excitation energies for radicals, encouraging the development of related seniority-based coupled cluster methods.
%\bigskip
%\begin{center}
%        \boxed{\includegraphics[keepaspectratio,width=2in]{TOC}}
%\end{center}
%\bigskip
\end{abstract}

@@

The most well-known CI route is based on the excitation degree.
Starting from a reference determinant, typically the Hartree-Fock (HF) determinant,
one generates all determinants that are connected to it by exciting at most $e$ electrons.
The excitation degree $e$ thus defines the order of the approximate excitation-based CI (eCI) model:
CI with single excitations (CIS); CI with single and double excitations (CISD), CI with single, double and triple excitations (CISDT), and so on.
The eCI route rather quickly captures dynamic (weak) correlation, though it struggles with the description of static (strong) correlation.

A different CI route is based on the seniority number $s$ (the number of unpaired electrons in a given determinant).
In seniority-based CI (sCI) \cite{Bytautas_2011,Allen_1962,Smith_1965,Veillard_1967}, there is no reference determinant, and one accounts for all determinants having seniority equal or less than $s$.
In contrast to eCI, sCI recovers static correlation more efficiently and works well to describe molecular dissociation, \cite{Bytautas_2015,Alcoba_2014,Alcoba_2014b}
at the expense of a poorer account of dynamic correlation and a higher computational cost.
For systems with an even number of electrons, the first approximate model is defined by $s=0$ (sCI0), usually referred to as doubly-occupied CI (DOCI),
which is followed by the higher-order models sCI2, sCI4, and so on.
For an odd number of electrons, the seniority route follows along odd numbers of $s$: sCI1, sCI3, and so on.

We have recently introduced a third CI route, hierarchy CI (hCI), \cite{Kossoski_2022}
where the Hilbert space is partitioned according to a hierarchy parameter $h$ that combines the excitation degree $e$ and the seniority number $s$ defined as $h = (e+s/2)/2$.
This definition ensures that all classes of determinants whose number share the same scaling with system size are included at the same hierarchy $h$.
For different properties and closed-shell molecular systems, 
hCI was found to display an overall faster convergence toward the FCI results than the traditional eCI route. \cite{Kossoski_2022}
In this sense, it was able to recover both static and dynamic correlation more effectively than either eCI or sCI.
However, hCI has only been defined for a closed-shell reference, thus being limited to the ground state of systems with an even number of electrons.

There are two possible approaches to target excited states with CI methods.
One can employ the ground-state HF orbitals and obtain excitation energies from the higher-lying eigenvalues of the CI matrix, which we refer to as ground-state based approach.
Instead, one may optimize the orbitals for the excited state of interest (described with an appropriate reference), followed by a separate CI calculation,
in a so-called state-specific approach ($\Delta$CI).
There has been a recent surge in the development of state-specific methods, covering 
single-reference and multiconfigurational self-consistent field,
\cite{Ziegler_1977,Burton_2021,Shea_2018,Tran_2019,Tran_2020,Hardikar_2020,Burton_2022,Hanscam_2022,Kossoski_2023,Marie_2023}
% references for SCF
%\cite{Ziegler_1977,Burton_2021}
% multiconfigurational SCF, \cite{Shea_2018,Tran_2019,Tran_2020,Hardikar_2020,Burton_2022,Kossoski_2023,Marie_2023}
density functional theory, 
\cite{Filatov_1999,Kowalczyk_2011,Kowalczyk_2013,Gilbert_2008,Barca_2018,Hait_2020,Hait_2021,Hardikar_2020,Zhao_2020,Levi_2020,Carter-Fenk_2020,Toffoli_2022,Schmerwitz_2022}
and coupled cluster (CC)
\cite{Piecuch_2000,Mayhall_2010,Lee_2019,Kossoski_2021,Marie_2021,Rishi_2023}
methods.
In particular, by employing a minimal configuration state function (CSF) reference, 
we have recently shown that excitation-based $\Delta$CI models deliver far more accurate excitation energies than the ground-state based analogues. \cite{Kossoski_2023}

Here, our first goal is to generalize hCI for an arbitrary type of reference, thus extending its applicability from ground-state closed-shells \cite{Kossoski_2022}
to radicals and state-specific excited state calculations.
This is done in Sec.~\ref{sec:hCI}.
The computational details about the implementation and the specific calculations performed here are presented in Sec.~\ref{sec:compdet}.
In Sec.~\ref{sec:res_A}, we assess the performance of hCI, eCI and sCI models by calculating various properties of four ground-state radicals, which comprises our second goal.
Our third goal is to evaluate how perturbation theory, more precisely the second-order Epstein-Nesbet (EN2) perturbative correction (both standard and renormalized), \cite{Garniron_2019}
impacts various properties of ground-state closed-shell systems and radicals, for hCI, eCI, and sCI models.
This part is presented in Sec.~\ref{sec:res_B}.
Inspired by the results of hCI for ground-state closed-shell systems, \cite{Kossoski_2022} 
and by the promising set of excitation-based $\Delta$CI, \cite{Kossoski_2023} here we explore hCI models for excited states, following both ground-state based and state-specific approaches.
In thise sense, our fourth goal is to assess the accuracy of hCI models for excited states, which is the subject of Sec.~\ref{sec:res_C}.
Furthermore, to the best of our knowledge, sCI models have not yet been directly used to target excited states,
despite the growing number of methods that exploit the concept of seniority number,
for both ground
\cite{Limacher_2013,Limacher_2014,Tecmer_2014,Boguslawski_2014a,Boguslawski_2015,Boguslawski_2014b,Boguslawski_2014c,Johnson_2020,Henderson_2014,Stein_2014,Henderson_2015,Chen_2015,Bytautas_2018,Marie_2021,Boguslawski_2021,Tecmer_2022,Mamache_2023} 
and excited states.
\cite{Boguslawski_2016b,Boguslawski_2016c,Boguslawski_2019,Nowak_2019,Kossoski_2021,Marie_2021,Tecmer_2022,Rishi_2023,Nowak_2023} 
Our fifth and last goal, detailed in Sec.~\ref{sec:res_D}, is therefore to define and gauge ground-state based and state-specific sCI models for excited states.
Sec.~\ref{sec:conclusion} closes the present contribution with the main conclusions and perspectives.

@@

We introduced hCI \cite{Kossoski_2022} as a particular truncation of the Hilbert space,
represented as a two-dimensional map of determinants built from their seniority $s$ and their excitation degree $e$ with respect to a reference determinant (see Fig.~\ref{fig:hCI}).
hCI was defined such that all classes of determinants whose number $\Ndet$ share the same scaling with the number of basis functions $N$ are included at the same order of truncation $h$,
which means moving diagonally in the seniority-excitation map (denoted by the color tones in Fig.~\ref{fig:hCI}).

@@

hCI was initially introduced for a closed-shell reference. \cite{Kossoski_2022}
Here, we generalize it for an arbitrary Slater determinant reference, including systems with an odd number of electrons.
With respect to a given reference determinant, we define the hierarchy $h$ of a candidate determinant to be included in the truncated hCI model as
\begin{equation}
  \label{eq:h}
  h = \frac{e+ (s-s_0)/2}{2},
\end{equation}
where $s$ and $s_0$ denote the seniority of the candidate and reference determinants, respectively, and $e$ represents the excitation degree that connects them.
The term $(s-s_0)/2$ is always an integer, with $s$ and $s_0$ being even (odd) for systems with an even (odd) numbers of electrons. The excitation degree $e$ is also an integer.
Therefore, $h$ can assume integer or half-integer values, for any type of reference determinant.
For a given hCI model defined by $h$, we include all the candidate determinants having a hierarchy less than or equal to $h$ with respect to any determinant in the reference.
This definition allow us to build multireference hCI models as well.
Notice that Eq.~\eqref{eq:h} simplifies to the previous definition \cite{Kossoski_2022} for the case of a closed-shell reference determinant, when $s_0 = 0$.
hCI can be viewed as the CI model that provides increasingly more dissimilar determinants from the given reference determinant.
The level of dissimilarity is represented by the hierarchy $h$, which accounts for differences in orbital occupation (through the excitation degree $e$)
and differences in the number of unpaired electrons (through the term $(s-s_0)/2$).

The definition in Eq.~\eqref{eq:h} guarantees that all classes of determinants sharing the same scaling of $\Ndet$ with $N$ enter at the same hierarchy $h$.
% computational scaling
We show in Fig.~\ref{fig:hCI} how the seniority-excitation map of determinants is partitioned for a closed-shell reference, and for open-shell references with one or two unpaired electrons.
These references are suitable for describing the ground state of a closed-shell system, of an open-shell system, and a singly-excited state of a closed-shell, respectively.
For these three types of reference, the classes of determinants included up to hCI1 are presented in Fig.~\ref{fig:determinants}.

%%% FIG 2 %%%
\begin{figure*}%[h!]
\includegraphics[width=1.0\linewidth]{determinants}
\caption{Classes of determinants generated in hCI0, hCI0.5 and hCI1, for three different reference determinants:
closed-shell (top left), open-shell with one unpaired electron (top right) and open-shell with two unpaired electrons (bottom).
Each class is determined by the specific seniority and occupation of the orbital blocks of the reference (doubly-occupied, singly-occupied and unoccupied).}
\label{fig:determinants}
\end{figure*}
%%% %%% %%%

The first order for a closed-shell reference determinant is hCI1 (see left of Fig.~\ref{fig:hCI}), which accounts for all single excitations (as CIS) 
plus all paired doubled excitations (two electrons promoted from the same occupied orbital into the same virtual orbital),
as shown in the top left of Fig.~\ref{fig:determinants}.
For both classes of determinants, their number scales as $N^2$, and are thus taken into account at the same hierarchy of hCI ($h=1$ in this case).
An odd number of excitations from a closed-shell reference always leave unpaired electrons, hence the empty blocks at $s=0$ for odd $e$.
hCI1.5 augments the set of hCI1 determinants with the set of double excitations where two electrons are unpaired.
At the next integer order, hCI2 incorporates all classes of determinants where $\Ndet$ scale as $N^4$.
In total, it accounts for all single and double excitations (as CISD), plus the subset of triple excitations that leave only two unpaired electrons, 
plus the subset of quadruple excitations where no electrons are unpaired.
More generally, in hCI, the number of determinants $\Ndet$ scales polynomially with the number of basis functions $N$ as $N^{2h}$.
Likewise, in eCI, $\Ndet$ scales as $N^{2e}$.
The computational scaling along the two routes is shown in Tab.~\ref{tab:scaling}.

\begin{table}[ht!]
\caption{Computational scaling in terms of the number of basis functions $N$ for the hierarchy-based and excitation-base CI routes, with and without the EN2 perturbative correction.}
\label{tab:scaling}
\begin{ruledtabular}
\begin{tabular}{ll|ll|l}
 \mc{2}{c|}{hierarchy-based}   &  \mc{2}{c|}{excitation-based} &  scaling \\
\hline
hCI1   &    -       & CIS    &   -       & $\order*{N^4}$ \\
hCI1.5 &    -       &  -     &   -       & $\order*{N^5}$ \\
\hline
hCI2   & hCI1+EN2   & CISD   & CIS+EN2   & $\order*{N^6}$ \\
hCI2.5 & hCI1.5+EN2 &  -     &   -       & $\order*{N^7}$ \\
\hline
hCI3   & hCI2+EN2   & CISDT  & CISD+EN2  & $\order*{N^8}$ \\
hCI3.5 & hCI2.5+EN2 &   -    &   -       & $\order*{N^9}$ \\
\hline
hCI4   & hCI3+EN2   & CISDTQ & CISDT+EN2 & $\order*{N^{10}}$ \\
\end{tabular}
\end{ruledtabular}
\end{table}

For an odd number of electrons, $s$ assumes odd values starting from $s=1$.
The simplest reference is an open-shell determinant with one unpaired electron (see center pannel of Fig.~\ref{fig:hCI} and top right of Fig.~\ref{fig:determinants}).
In contrast to the closed-shell case, here there are no empty blocks in the excitation-seniority map.
Actually, the hCI series displays the hCI0.5 level, accounting only for the single excitations from and into the singly-occupied orbital.
For HF orbitals, however, these excitations do not connect with the reference, and hCI0.5 provides the same energy as restricted open-shell HF.
More generally, since there are more types of orbitals for the open-shell reference (doubly-occupied, singly-occupied and unoccupied) than for the closed-shell reference (doubly-occupied or unoccupied),
there are correspondingly more classes of determinants at a given level of hCI.
This can be appreciated from the comparison of hCI1 for closed- and open-shell references, shown in Fig.~\ref{fig:determinants}.

Finally, one can employ a high-spin reference determinant to define the hCI models.
The case of two unpaired electrons is shown in the right pannel of Fig.~\ref{fig:hCI},
which would be appropriate for singly-excited states of closed-shells.
The key difference in the hCI sequence, with respect to the closed-shell case, lies in the displacement by one block to the right in the seniority-excitation map,
reflecting the shift from the $s_0=0$ to the $s_0=2$ reference.
In the two cases discussed, the hCI0 level only included the reference determinant.
For a $s=2$ reference, however, hCI0 not only accounts for the reference determinant,
but also for the two closed-shell determinants produced by the single excitation that pairs the two unpaired electrons.
The hCI1 classes of determinants for the $s_0=2$ reference can be seen in the bottom of Fig.~\ref{fig:determinants},
which clearly outnumber the fewer classes associated with the $s_0=0$ and $s_0=1$ references.
One could adopt even higher-spin references, further displacing the hCI sequence to the right in the seniority-excitation map.
In this case, hCI0 would include all the lower spin determinants obtained by partially or totally pairing the unpaired electrons.
Notice that the number of such determinants do not depend on the size of the system (scaling as $N^0$), and are thus included at the $h=0$ level, in line with the spirit of hCI.
In a similar fashion, hCI can be built on top of a high-spin reference for odd number of electrons.

@@

The hCI models introduced here were implemented in {\QP} \cite{Garniron_2019} through a straightforward modification of the
\textit{configuration interaction using a perturbative selection made iteratively} (CIPSI) algorithm. \cite{Huron_1973,Giner_2013,Giner_2015,Garniron_2018}
By allowing only for the determinants that are connected with the reference determinant(s) up to a given maximum hierarchy $h$,
the CIPSI algorithm is restricted to the truncated Hilbert space specified by the reference determinant(s) and the value of $h$.
{\QP} \cite{Garniron_2019} was also employed to perform all the present eCI, sCI and FCI calculations.
In a given calculation, the energies are considered to be converged when the (largest) EN2 correction computed in the truncated Hilbert space 
lies below \SI{0.01}{\milli\hartree}. \cite{Garniron_2018}
This selected CI procedure requires considerably fewer determinants than the total number of determinants in the truncated Hilbert space,
while delivering fairly converged energies.
The ground- and excited-state CI energies are obtained with the Davidson iterative algorithm. \cite{Davidson_1975}
For a given approximate CI model, we further evaluated the standard and the renormalized EN2 perturbative correction. \cite{Garniron_2019} 
This calculation involves a single loop over the determinants left outside the truncated (internal) space but connected to it via at most double excitations.
Looping over these external doubly excited determinants has a computational scaling equal to $\Ndet$,
thus $\order*{N^{2e+4}}$ for eCI and $\order*{N^{2h+4}}$ for hCI, where $e$ or $h$ define the internal CI space.
For example, hCI2 and CISD present a $\order*{N^6}$ computational scaling, whereas hCI2+EN2 and CISD+EN2 scale as $\order*{N^8}$,
though with a small prefactor stemming from the EN2 calculation.
The computational scaling associated with the CI+EN2 calculations is also presented in Tab.~\ref{tab:scaling}.

To gauge the performance of hCI, eCI and sCI for radical systems,
we have calculated the ground-state potential energy curves (PECs) for the dissociation of four radicals:
\ce{OH}, \ce{CN}, vinyl (\ce{C2H3}), and \ce{H7}.
The CI calculations employed the ground-state restricted open-shell HF orbitals, described with the cc-pVDZ basis set, and within the frozen core approximation.
For such small systems and basis set, FCI is attainable and provides the reference results for gauging the approximate CI models.
The equilibrium geometry of vinyl was taken from Ref.~\onlinecite{Loos_2020}, and is also reproduced in the {\SupInf}.
Their PECs were computed along the \ce{C=C} double bond breaking coordinate, with the remaining internal coordinates kept frozen.
For \ce{H7}, we considered equally spaced and linearly arranged hydrogen atoms, and the PECs were computed along the symmetric dissociation coordinate.

The results were analyzed along the same lines of our previous report on hCI for closed-shell systems. \cite{Kossoski_2022}
Namely, for the different CI models considered here, 
we evaluated the convergence of the non-parallelity error (NPE), the distance error, the harmonic vibrational frequencies, and the equilibrium bond lengths, as functions of $\Ndet$.
The NPE of a given level of theory is defined as the maximum minus the minimum differences between its corresponding PEC and the FCI PEC, for a given range of coordinates.
Here, we redefine the previous definition of the distance error \cite{Kossoski_2022} to accommodate for the fact the approximate PEC might appear below the FCI one when perturbative corrections are employed.
In such cases and with the previous definition, undesired negative values could be attained.
The distance error is redefined based on the signed differences between two PECs, as the absolute value of their maximum difference plus the absolute value of their minimum difference, 
evaluated at given coordinate interval.
This new definition measures how close are two PECs, remaining always non-negative.
From here on, we employ equilibrium properties when referring to both the equilibrium geometry and the harmonic vibrational frequency.
Details about how the equilibrium properties were obtained from the calculated PECs,
along with the ranges defining the NPE and distance errors can be found in the \SupInf.

The various CI models introduced here were further assessed based on calculated vertical excitation energies for 60 electronic states,
from 18 closed-shell systems and 8 radicals, with geometries extracted from the QUEST database. \cite{Veril_2021}
The full set of excited states and calculated excitation energies, for the various CI models, are provided in the {\SupInf}.
We employed the aug-cc-pVDZ basis set for systems having up to three non-hydrogen atoms and the 6-31+G(d) basis set for the larger ones.
Core orbitals were frozen.
For the radicals, we further impose the CI solutions to be eigenstates of the spin angular momentum operator, which implies in accounting for a set of appropriate spin-flipped determinants.
In this sense, our CIS calculations for radicals actually correspond to the so-called extended CIS, \cite{Maurice_1996} and equivalently for the other hCI, eCI and sCI models.
We performed calculations following both the standard ground-state based CI route and the state-specific CI route. \cite{Kossoski_2023}
For the latter, we employed the state-specific orbitals obtained in Ref.~\onlinecite{Kossoski_2023}.
Notice that, in contrast to eCI, the energies obtained with sCI and hCI models are not invariant under rotations within the occupied and virtual subspaces.
The computed excitation energies were benchmarked against the reference values of FCI quality provided in the QUEST database. \cite{Veril_2021}
